---
title: "Module 5 Assignment"
author: "Daniel Bihnam"
format: docx
editor: visual
---

## Load in and Process Data

```{r}
load('C:/Users/m293215/Downloads/tobacco_clr.Rdata')

data10 <- tobacco_clr$data[,1:10]
h10 <- tobacco_clr$H[,1:10]
color10 <- tobacco_clr$sample.color[1:10]
pch10 <- tobacco_clr$sample.pch[1:10]
otu10 <- tobacco_clr$otu.names[1:10]
```

First, I loaded in my data. Then, in order to specify the first 10 entries of the data, I used \[1:10\] and \[,1:10\] for normal vectors and matrix vectors.

## Network with ppcor and igraph (Partial Correlation Analysis)

```{r}
library(ppcor)
round(pcor(data10)$p.value,3)

library(igraph)
network <- (pcor(data10)$p.value<=0.05)*1

diag(network) <- 0
network.plot <- graph_from_adjacency_matrix(network)
#Create the plot
V(network.plot)$label <- otu10
#Use the first 10 OTUs to label the nodes
plot(network.plot,edge.arrow.size=.5,main='DB: 10 OTU Network 1')

```

Using ppcor and igraph like we were shown in lecture, I created a network using a Gaussian graphical model. This model is able to display the direction of relationships between our first 10 OTUs in the dataset. We can see that there is a lot of relationship surrounding cowanii, staphylococcus, chromatiacaea, and enterobacteriaceae. We can also see equorum and salaria are only related with each other, and not the other 8 OTUs. This network is the main network required for this assignment.

## Network with glasso and qgraph (Graphical Lasso)

```{r}
library(glasso)

cm <- cor(data10)
#Create correlation matrix
pcm <- glasso::glasso(cm,rho=0.1)
#Create partial correlation matrix using glasso library

library(qgraph)
gauss <- qgraph(pcm,layout='spring',labels=otu10,title='DB: 10 OTU Network 2')
#Create plot
```

While working on this assignment, I came across another package called glasso. Glasso reminded me of the lasso model that was used in our yellowdig discussion this week, so I decided to experiment with it and see how it differs from our previous model. It can be used to infer conditional dependencies between variables. This model tells a different story than the previous one, as we can see some bacteria such as equorum having a lot of dependencies on and for other OTUs. Whereas in the first model, it was seperate from most of these OTUs. This was not required for this assignment, but I found it interesting and wanted to include it as an addition.
