---
title: "LSC541 Module 2: Assignment"
author: "Daniel Bihnam"
format: docx
editor: visual
---

## Read in Data

```{r}
getwd()
setwd("C:/Users/Daniel/Documents")
autism0 <- read.table('C:/Users/Daniel/Downloads/data1_LSC598.txt', header = T)
autism <- na.omit(autism0)
#Reading in data and omitting a N/A value from the data
```

## Qualitative Data

```{r}
library(ggplot2)

ggplot(autism,aes(x=gender,fill=gender))+
  geom_bar()+
  xlab('Gender')+
  ylab('Count')+
  ggtitle('DB: Females vs Males in Autism Study')

ggplot(autism,aes(x=as.factor(group),fill=as.factor(group)))+
  geom_bar()+
  xlab('Group')+
  ylab('Count')+
  ggtitle('DB: Comparing Participants with and Without Autism')+
  scale_fill_discrete(labels=c('No Autism','Autism'))

ggplot(autism,aes(x=as.factor(group),fill=gender))+
  geom_bar()+
  xlab('Group')+
  ylab('Count')+
  ggtitle('DB: Comparing Participant Gender with Autism')+
  scale_x_discrete(labels=c('0'='No Autism','1'='Autism'))
```

1.  Barplot of the count of females and males in the data set. There are more than twice the amount of females in the data set as there are males.

2.  Barplot comparing amount of people with and without autism in this study. There are more people included in this study that have autism than those that do not have autism.

3.  These barplots can be combined to display both variables at the same time. Visually, there looks to be a similar proportion of males and females with and without autism. It is important to note that this data set may not be a strong predictor for a general population. The data is made up predominantly of females, and the prevalence of autism among the participants is higher than that of the general population.

## Quantitative Data

```{r}
ggplot(autism,aes(x=as.factor(0),y=age_month))+
  geom_boxplot(width=0.2)+
  ylab('Age in Months')+
  xlab('All Participants')+
  ggtitle('DB: Age in Months of all Participants')


ggplot(autism,aes(x=as.factor(0),y=vitD_level))+
  geom_boxplot(width=0.2)+
  ylab('Vitamin D Level')+
  xlab('All Participants')+
  ggtitle('DB: Vitamin D Levels of all Participants')

```

1.  Boxplot of participant ages in months. There was no second variable listed for us to compare to, so I just plotted the age variable against 0 as a placeholder. This data is slightly skewed towards younger participants.

2.  Boxplot of participant vitamin D levels. Similar to the previous plot, there was no variable we were told to compare to, so I just plotted it against x=0 as a placeholder. This data is slightly skewed towards those with lower vitamin D levels, but we can see many outliers with very high vitamin D levels.

## Hypothesis Testing of Continuous Variables

```{r}
t.test(autism$vitD_level)

t.test(autism$age_month)
```

1.  One-sample t-test for the vitD_level variable. Our sample mean was found to be equal to 35.96, and our t-test yielded a significant p-value (\<0.05). We can report with 95% confidence that the true mean vitamin D level lies between 32.63 and 39.30.

2.  One-sample t-test for the age in months variable. Our sample mean was found to be equal to 51.07, and our t-test yielded a significant p-value (\<0.05). We can report with 95% confidence that the true mean age in months lies between 48.73 and 53.41. It is important to note that age was grouped into full months in the collection of this data. However, age is still a continuous variable even though it may appear to be discrete.
